Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-02-09 paper/paper.tex
Update Abstract

In this paper, we present a novel method to characterize OCs.
Our method employs a model built on \emph{Artificial Neural Networks} (ANNs).
More specifically, we adapted a state of the art model, the
\emph{Deep Embedded Clustering} (DEC) model for our purpose.

The developed method aims to improve classical state of the arts techniques.
We improved not only in terms of computational efficiency, (with lower
computational requirements), but in usability (reducing the number
of hyperparameters to get a good characterization of the analyzed clusters).
For our experiments, we used the \emph{Gaia DR2 database} as the data source,
and compared our model with the clustering technique \emph{K-Means}.

Our method achieves good results,
becoming even better (in some of the cases) than current techniques.
}

>>>>

In this paper, we present a novel method to characterize OCs. Our method employs a
model built on \emph{Artificial Neural Networks} (ANNs). More specifically, we adapted
a state of the art model, the \emph{Deep Embedded Clustering} (DEC) model for our purpose.
The developed method aims to improve classical state of the arts techniques. We improved
not only in terms of computational efficiency (with lower computational requirements),
but in usability (reducing the number of hyperparameters to get a good characterization
of the analyzed clusters). For our experiments, we used the \emph{Gaia DR2 database} as
the data source, and compared our model with the clustering technique \emph{K-Means}. Our
method achieves good results, becoming even better (in some of the cases) than current techniques.
}

### 2021-02-09 thesis/thesis.tex
thesis: Minor corrections
thesis: Fix appending paper
Update Abstract

The characterization and understanding of \emph{Open Clusters} allow us to better understand properties and mechanisms about the Universe
such as stellar formation and the kind of regions where these events occur.
They also provide information about stellar processes and the evolution of the galactic disk.

In this work, we develop a method to characterize open clusters by using \emph{artificial intelligence} tools,
such as clustering by \emph{K-Means} and clustering based on \emph{Artificial Neural Networks} by implementing the \emph{Deep Embedded Clustering} model.
We are using \emph{Gaia DR2 database} as data source for testing our models.

The developed method aims to improve current techniques of the state of the art.
We achieve improvements not only in terms of \emph{computational efficiency}, with lower computational requirements, but in \emph{usability},
reducing the number of hyperparameters to obtain a good characterization of the analyzed clusters.

Our method achieves good results, becoming even better in some cases when the results are compared with current techniques.

@@

En este trabajo se desarrolla un método para caracterizar estos cúmulos mediante el uso de herramientas de \emph{inteligencia artificial}
como agrupamiento por \emph{K-Medias} o agrupamiento basado en \emph{Redes Neuronales Artificiales} mediante la implementación del
modelo \emph{Deep Embedded Clustering} aplicados sobre conjuntos de datos extraídos de la base de datos de \emph{Gaia DR2}.

El método desarrollado pretende mejorar los ya existentes tanto en términos de \emph{eficiencia computacional}, con menores requisitos de cálculo,
como en \emph{facilidad de uso}, reduciendo el número de hiperparámetros a configurar para obtener una buena caracterización de los cúmulos analizados.

Como se detalla más adelante, nuestro método consigue buenos resultados llegando a ser incluso mejor en algunos casos cuando se comparan
los resultados con otras técnicas actuales.

@@

\includepdf[pages={-},noautoscale]{../paper/paper.pdf}

>>>>

The characterization and understanding of \emph{Open Clusters} (OCs) allow us to
understand better properties and mechanisms about the Universe such as stellar
formation and the regions where these events occur. They also provide information
about stellar processes and the evolution of the galactic disk.

In this work, we present a novel method to characterize OCs. Our method employs a
model built on \emph{Artificial Neural Networks} (ANNs). More specifically, we adapted
a state of the art model, the \emph{Deep Embedded Clustering} (DEC) model for our purpose.
The developed method aims to improve classical state of the arts techniques. We improved
not only in terms of computational efficiency (with lower computational requirements),
but in usability (reducing the number of hyperparameters to get a good characterization
of the analyzed clusters). For our experiments, we used the \emph{Gaia DR2 database} as
the data source, and compared our model with the clustering technique \emph{K-Means}. Our
method achieves good results, becoming even better (in some of the cases) than current techniques.

@@

En este trabajo presentamos un método novedoso para caracterizar estos cúmulos.
Nuestro método hace uso de un modelo basado en \emph{Redes Neuronales Artificiales}.
Más concretamente, adaptamos un modelo del estado del arte, el \emph{Deep Embedded
Clustering} (DEC), a nuestro problema. El método desarrollado tiene como objetivo
mejorar las técnicas clásicas del estado del arte. Con nuestro método, no sólo
mejoramos en términos de eficiencia de cálculo (consiguiendo menores requisitos
computacionales), también mejoramos en usabilidad (reduciendo el número de
hiperparámetros para conseguir una buena caracterización de los cúmulos analizados).
Para nuestro experimentos, usamos la base de datos \emph{Gaia DR2} como fuente de datos,
y comparamos nuestro modelos con el algoritmo de clustering \emph{K-Medias}. Nuestro
método consigue buenos resultados, siendo incluso mejor (en algunos casos) que las
técnicas actuales.

@@

\setboolean{@twoside}{false}
\addtocontents{toc}{\protect\renewcommand{\protect\numberline}[1]{}}
\includepdf[pages=-, offset=75 -75, addtotoc={1, chapter, 1, Paper, chap:paper}]{../paper/paper.pdf}
